\MakeUppercase{Graph Product Structure for Non-Minor-Closed Classes}

### 2022-03-08 k-planar.tex
changes

Our proof of \cref{1-planar} follows quickly from the following technical lemma, which is an extension of the analogous result for plane graphs \cite{DJMMUW20}.
\begin{lem}
	\label{induction} The setup:
	\begin{compactenum}
		\item Let $G$ and $G'$ be defined as above.
		\item Let $T$ be a BFS spanning tree of $G'$ rooted at some vertex $r$.
		\item For every integer $j\ge 0$, let $L_j=\{v\in V(G):\dist_T(r,v)=j\}$.
		\item Let $F$ be a cycle in $G'$ with $r$ in the exterior of $F$ and such that
		\begin{compactenum}
			\item No edge of $F$ is crossed by any edge of $G$; and
			\item $V(F)$ can be partitioned into $P_1,\ldots,P_k$, for some $k\in\{1,2,3\}$ such that for each $i\in\{1,\ldots,k\}$,
			\begin{compactenum}
				\item $F[P_i]$ is a path; and
				\item $|V(P_i)\cap L_j| \le 15$ for every integer $j\ge 0$.
			\end{compactenum}
		\end{compactenum}
	\note{DW}{Should the vertices already grabbed by $P_1,P_2,P_3$ be mentioned in the setup? The new paths $Q_1,Q_2,Q_3$ need to avoid the vertices already grabbed by $P_1,P_2,P_3$. Basically, I am concerned about what happens when distinct vertical paths in $T$ want to grab the same vertex.}
  \note{PM}{No, at this point we're only promising an $H$-partition of $N$, the part of $G$ contained in the uncrossed cycle $F$.}
		\item Let $N$ and $N'$ be the subgraphs of $G$ and $G'$ consisting only of those edges and vertices contained in $F$ or the interior of $F$.
	\end{compactenum}
	Then $N$ has an $H$-partition $\PP=\{S_x : x\in V(H)\}$ such that:
	\begin{compactenum}
		\item $H$ is planar;
		\item for every integer $j\ge 0$ and every $x\in V(H)$, $|S_x\cap L_j|\le 15$;
		\item for each $i\in\{1,\ldots,k\}$, there exists some $x_i\in V(H)$ such that $P_i=S_{x_i}$
		\note{DW}{Should this be $P_i\subseteq S_{x_i}$? (since the path grabs other vertices into its part)}\note{PM}{No, for the same reason as above.}; and
		\item $H$ has a tree-decomposition in which every bag has size at most 4 and such that some bag contains $x_1,\ldots,x_k$.
	\end{compactenum}
\end{lem}

@@

\begin{proof}
	For a vertex $x$ of $G_0$ with degree at most $\Delta$, and for $i\in\{1,\dots,k-1\}$, say a vertex $v$ is \defin{$i$-close} to $x$ if there is a $vx$-path $P$ in $G_0$ of length at most $i$ such that every internal vertex in $P$  has degree at most $\Delta$ in $G_0$.
	For each edge $vw$ of $G$, say that $vw$ \defin{passes through} each internal vertex on $P_{vw}$.	Say $vw$ passes through $x$. Then $v$ is $i$-close to $x$ and $w$ is $j$-close to $x$ for some $i,j\in\{1,\dots,k-1\}$ with $i+j\leq k$. At most $\Delta^{i}$ vertices are $i$-close to $x$.
	Thus, the number of edges of $G$ that pass through $x$ is at most
	\[
	\sum_{i=1}^{k-1} \sum_{j=1}^{k-i} \Delta^i \Delta^j
	= \sum_{i=1}^{k-1} \Delta^i  \sum_{j=1}^{k-i} \Delta^j
	< \sum_{i=1}^{k-1} \Delta^i  2 \Delta^{k-i}
	= \sum_{i=1}^{k-1} 2\Delta^k
	< 2k \Delta^k \enspace.
	\]
	Fix a $(\Sigma,k)$-plane drawing of $G_0$ in a surface $\Sigma$ of Euler genus at most $g$. Choose $\epsilon>\epsilon'>0$.
	For each vertex $v$ of $G_0$, let $B_v$ be the \defin{ball} $\{x\in \mathbb{R}^2:\dist_(x,v)\leq\epsilon\}$
	For each edge $vw$ of $G_0$, let $C_{vw}$ be the \defin{channel} $\{x\in \mathbb{R}^2:\dist(x,vw)\leq\epsilon'\}\setminus(B_v\cup B_w)$.
	We may choose $\epsilon>\epsilon'>0$ sufficiently small so that
	(i) $B_v\cap B_w=\emptyset$ for distinct $v,w\in V(G_0)$,
	(ii) $C_{vw}\neq\emptyset$ for $vw\in E(G_0)$,
	(iii) $B_v\cap C_{xy}=\emptyset$ for $v\in V(G_0)$ and $xy\in E(G_0)$, and
	(iv) $C_{vw}\cap C_{xy}=\emptyset$ for distinct $vw,xy\in E(G_0)$.
	Draw each edge $vw$ of $G$ following the sequence of balls and channels defined by the vertices and edges in $P_{vw}$.
	This can be done so that whenever edges $vw$ and $xy$ of $G$ cross, the crossing point is in $B_z$ for some vertex $z$ of $G_0$ that is internal in $P_{vw}$ or in $P_{xy}$, and $vw$ and $xy$ cross at most once in each such $B_z$.

@@

Prior to the current work, the strongest upper bound on the non-repetitive chromatic number of $n$-vertex  $k$-planar graphs was $O(k\log n)$ \cite{dujmovic.morin.ea:layered}.
=======
%      \includegraphics{figs/one_planar_example-1} &
%      \includegraphics{figs/one_planar_example-2} \\
%      (a) & (b) \\[1em]
%  \caption{Examples of (a)~a $1$-plane graph that is not edge maximal; (b)~an edge-maximal $1$-plane graph $G$; (c)~the kites in $G$; (d)~a plane multigraph $G'$ obtained by removing one spar from each kite of $G$.}

To understand the structure of $G$, it is helpful to consider the plane graph $G_0$ obtained by replacing each pair of edges $vw$ and $xy$ involved in a crossing $(p,vw,xy)$ with four edges $vp$, $wp$, $xp$, and $yp$ meeting at a newly added \defin{dummy vertex} $p$. Let $F$ be a face of $G_0$ and let $W:=w_0,\ldots,w_r$ be the facial walk around $F$. By edge maximality, if $W$ contains only non-dummy vertices, then it contains exactly three vertices and $F$ is bounded by three edges of $G$.
>>>>>>> 432123323ba65fbac04503f0de5e50ec4ebd89ce

If $W$ contains a dummy vertex $w_i$ then neither $w_{i-1}$ nor $w_{i+1}$ is a dummy vertex since if $w_{i-1}$ (respectively $w_{i+1}$) were a dummy vertex then the edge of $G$ that contains $w_{i-1}w_i$ (respectively $w_iw_{i+1}$) woul

@@

\begin{thm}
\label{StringPartition}
For integers $g\geq 0$ and $\delta\geq 2$, let $\ell:= \max\{2g,3\} \,(\delta^4 + 4 \delta^3 + 9 \delta^2 + 10 \delta + 4)$,
and $t:= \binom{ \delta+4}{3}-1$.
Then  every $(g,\delta)$-string graph:
\begin{compactitem}
	\item is contained in $H\boxtimes P \boxtimes K_{\ell}$ for some path $P$ and for some graph $H$ with treewidth $t$,
	\item has queue-number $\qn(G^k) \leq 3 \ell \, 2^t - \ceil{\tfrac{3}{2}\ell}$.
\end{compactitem}
\end{thm}

@@

In this section, we show that $k$-nearest neighbour graphs of point sets in the plane are $O(k^2)$-planar.  For two points $x,y\in\R^2$, let $d_2(x,y)$ denote the Euclidean distance between $x$ and $y$. The $k$-nearest-neighbour graph of a point set $P\subset\R^2$ is the geometric graph $G$ with vertex set $V(G)=P$, where the edge set is defined as follows. For each point $v\in P$, let $N_k(v)$ be the set of $k$ points i
=======
%      \includegraphics{figs/one_planar_example-1} &
%      \includegraphics{figs/one_planar_example-2} \\
%      (a) & (b) \\[1em]
%  \caption{Examples of (a)~a $1$-plane graph that is not edge maximal; (b)~an edge-maximal $1$-plane graph $G$; (c)~the kites in $G$; (d)~a plane multigraph $G'$ obtained by removing one spar from each kite of $G$.}

To understand the structure of $G$, it is helpful to consider the plane graph $G_0$ obtained by replacing each pair of edges $vw$ and $xy$ involved in a crossing $(p,vw,xy)$ with four edges $vp$, $wp$, $xp$, and $yp$ meeting at a newly added \defin{dummy vertex} $p$. Let $F$ be a face of $G_0$ and let $W:=w_0,\ldots,w_r$ be the facial walk around $F$. By edge maximality, if $W$ contains only non-dummy vertices, then it contains exactly three vertices and $F$ is bounded by three edges of $G$.
>>>>>>> 432123323ba65fbac04503f0de5e50ec4ebd89ce

If $W$ contains a dummy vertex $w_i$ then neither $w_{i-1}$ nor $w_{i+1}$ is a dummy vertex since if $w_{i-1}$ (respectively $w_{i+1}$) were a dummy vertex then the edge of $G$ that contains $w_{i-1}w_i$ (respectively $w_iw_{i+1}$) wouln $P$ closest to $v$. Then $vw\in E(G)$ if and only if $w\in N_k(v)$ or $v\in N_k(w)$. (The edges of $G$ are straight-line segments joining their endpoints.) See \citep{ProximityGraphs} for a survey of results on $k$-nearest neighbour graphs and other related proximity graphs.

>>>>

Our proof of \cref{1-planar} follows quickly from the following technical lemma, which is an extension of the analogous result for plane graphs \cite{DJMMUW20}.
\begin{lem}
	\label{induction} The setup:
	\begin{compactenum}
		\item Let $G$ and $G'$ be defined as above.
		\item Let $T$ be a BFS spanning tree of $G'$ rooted at some vertex $r$.
		\item For every integer $j\ge 0$, let $L_j=\{v\in V(G):\dist_T(r,v)=j\}$.
		\item Let $F$ be a cycle in $G'$ with $r$ in the exterior of $F$ and such that
		\begin{compactenum}
			\item No edge of $F$ is crossed by any edge of $G$; and
			\item $V(F)$ can be partitioned into $P_1,\ldots,P_k$, for some $k\in\{1,2,3\}$ such that for each $i\in\{1,\ldots,k\}$,
			\begin{compactenum}
				\item $F[P_i]$ is a path; and
				\item $|V(P_i)\cap L_j| \le 15$ for every integer $j\ge 0$.
			\end{compactenum}
		\end{compactenum}
% \note{DW}{Should the vertices already grabbed by $P_1,P_2,P_3$ be mentioned in the setup? The new paths $Q_1,Q_2,Q_3$ need to avoid the vertices already grabbed by $P_1,P_2,P_3$. Basically, I am concerned about what happens when distinct vertical paths in $T$ want to grab the same vertex.} \note{PM}{No, at this point we're only promising an $H$-partition of $N$, the part of $G$ contained in the uncrossed cycle $F$.}
		\item Let $N$ and $N'$ be the subgraphs of $G$ and $G'$ consisting only of those edges and vertices contained in $F$ or the interior of $F$.
	\end{compactenum}
	Then $N$ has an $H$-partition $\PP=\{S_x : x\in V(H)\}$ such that:
	\begin{compactenum}
		\item $H$ is planar;
		\item for every integer $j\ge 0$ and every $x\in V(H)$, $|S_x\cap L_j|\le 15$;
		\item for each $i\in\{1,\ldots,k\}$, there exists some $x_i\in V(H)$ such that $P_i=S_{x_i}$; 
%		\note{DW}{Should this be $P_i\subseteq S_{x_i}$? (since the path grabs other vertices into its part)}\note{PM}{No, for the same reason as above.}
 and
		\item $H$ has a tree-decomposition in which every bag has size at most 4 and such that some bag contains $x_1,\ldots,x_k$.
	\end{compactenum}
\end{lem}

@@

\begin{proof}
	For a vertex $x$ of $G_0$ with degree at most $\Delta$, and for $i\in\{1,\dots,k-1\}$, say a vertex $v$ is \defin{$i$-close} to $x$ if there is a $vx$-path $P$ in $G_0$ of length at most $i$ such that every internal vertex in $P$  has degree at most $\Delta$ in $G_0$.
	For each edge $vw$ of $G$, say that $vw$ \defin{passes through} each internal vertex on $P_{vw}$.	Say $vw$ passes through $x$. Then $v$ is $i$-close to $x$ and $w$ is $j$-close to $x$ for some $i,j\in\{1,\dots,k-1\}$ with $i+j\leq k$. At most $\Delta^{i}$ vertices are $i$-close to $x$.
	Thus, the number of edges of $G$ that pass through $x$ is at most
	\[
	\sum_{i=1}^{k-1} \sum_{j=1}^{k-i} \Delta^i \Delta^j
	= \sum_{i=1}^{k-1} \Delta^i  \sum_{j=1}^{k-i} \Delta^j
	< \sum_{i=1}^{k-1} \Delta^i  2 \Delta^{k-i}
	= \sum_{i=1}^{k-1} 2\Delta^k
	< 2k \Delta^k \enspace.
	\]
	Fix a $(\Sigma,k)$-plane drawing of $G_0$ in a surface $\Sigma$ of Euler genus at most $g$. Choose $\epsilon>\epsilon'>0$.
	For each vertex $v$ of $G_0$, let $B_v$ be the \defin{ball} $\{x\in \mathbb{R}^2:\dist(x,v)\leq\epsilon\}$
	For each edge $vw$ of $G_0$, let $C_{vw}$ be the \defin{channel} $\{x\in \mathbb{R}^2:\dist(x,vw)\leq\epsilon'\}\setminus(B_v\cup B_w)$.
	We may choose $\epsilon>\epsilon'>0$ sufficiently small so that
	(i) $B_v\cap B_w=\emptyset$ for distinct $v,w\in V(G_0)$,
	(ii) $C_{vw}\neq\emptyset$ for $vw\in E(G_0)$,
	(iii) $B_v\cap C_{xy}=\emptyset$ for $v\in V(G_0)$ and $xy\in E(G_0)$, and
	(iv) $C_{vw}\cap C_{xy}=\emptyset$ for distinct $vw,xy\in E(G_0)$.
	Draw each edge $vw$ of $G$ following the sequence of balls and channels defined by the vertices and edges in $P_{vw}$.
	This can be done so that whenever edges $vw$ and $xy$ of $G$ cross, the crossing point is in $B_z$ for some vertex $z$ of $G_0$ that is internal in $P_{vw}$ or in $P_{xy}$, and $vw$ and $xy$ cross at most once in each such $B_z$.

@@

Prior to the current work, the strongest upper bound on the non-repetitive chromatic number of $n$-vertex  $k$-planar graphs was $O(k\log n)$ \cite{dujmovic.morin.ea:layered}.

@@

\begin{thm}
\label{StringPartition}
For integers $g\geq 0$ and $\delta\geq 2$, let $\ell:= \max\{2g,3\} \,(\delta^4 + 4 \delta^3 + 9 \delta^2 + 10 \delta + 4)$ and $t:= \binom{ \delta+4}{3}-1$.
Then  every $(g,\delta)$-string graph:
\begin{compactitem}
	\item is contained in $H\boxtimes P \boxtimes K_{\ell}$ for some path $P$ and for some graph $H$ with treewidth $t$,
	\item has queue-number $\qn(G^k) \leq 3 \ell \, 2^t - \ceil{\tfrac{3}{2}\ell}$.
\end{compactitem}
\end{thm}

@@

In this section, we show that $k$-nearest neighbour graphs of point sets in the plane are $O(k^2)$-planar.  For two points $x,y\in\R^2$, let $d_2(x,y)$ denote the Euclidean distance between $x$ and $y$. The $k$-nearest-neighbour graph of a point set $P\subset\R^2$ is the geometric graph $G$ with vertex set $V(G)=P$, where the edge set is defined as follows. For each point $v\in P$, let $N_k(v)$ be the set of $k$ points in $P$ closest to $v$. Then $vw\in E(G)$ if and only if $w\in N_k(v)$ or $v\in N_k(w)$. (The edges of $G$ are straight-line segments joining their endpoints.) See \citep{ProximityGraphs} for a survey of results on $k$-nearest neighbour graphs and other related proximity graphs.
